---
title: "Figure S3 - MAPS across tissue activity groups"
author: "Patrick Short"
date: "19 January 2017"
output: html_document
---

```{r load the unaff parental variants}
source("~/phd/code/dddMAPS/dddMAPS/MAPS.R")
source("~/phd/code/dddMAPS/dddMAPS/MAPS_plotting_extras.R")
source("../R/annotation_tools.R")
library(stringr)

unaff_parent_variants = read.table("../data/unaffected_parent_alleles_all_chromosomes.FS_20.phylop.txt", header = TRUE, sep = "\t")

unaff_parents_dhs = read.table("../data/unaffected_parent_alleles_all_chromosomes.CADD.DHS_broad_peak_fdr_0.01.chromHMM.AllRoadmapTissues.txt", header = TRUE, sep = "\t")

unaff_parent_variants = merge(unaff_parent_variants, unaff_parents_dhs)


```

First, let's use the dddMAPS package to look for evidence of purifying selection across the three different element sets.

```{r load the elements of interest}

noncoding_control_elements = read.table("../data/noncoding_control_elements.10bp_buffer.min10_coverage.30bp_element_minimum.30x_probe_coverage_minimum.no_ddg2p_overlap.txt", header = TRUE, sep = "\t")
coding_elements = read.table("../data/gencode.v19.CDS.probe_overlap.min10_coverage.txt", header = TRUE, sep = "\t")
heart_elements = read.table("../data/heart_elements.10bp_buffer.min10_coverage.30bp_element_minimum.txt", header = TRUE, sep = "\t")
conserved_elements = read.table("../data/conserved_elements.min10_coverage.txt", header = TRUE, sep = "\t")
enhancer_elements = read.table("../data/enhancer_elements.min10_coverage.txt", header = TRUE, sep = "\t")


```

Now, we can test evidence for purifying selection across different tissue sets in the unaffected parents.

```{r define the tissue sets}

ESC = c("E002", "E008", "E001", "E015", "E014", "E016", "E003", "E024")
iPSC = c("E020", "E019", "E018", "E021", "E022")
ES_derived_tissues = c("E007", "E009", "E010", "E013", "E012", "E011", "E004", "E005", "E006")
blood = c("E062", "E034", "E045", "E033", "E044", "E043", "E039", "E041", "E042", "E040", "E037", "E048", "E038", "E047")
hsc_and_b_cell = c("E029", "E031", "E035", "E051", "E050", "E036", "E032", "E046", "E030")
mesench = c("E026", "E049", "E025", "E023")
myostat = c("E052")
epithelial = c("E055","E056","E059","E061","E057","E058","E028","E027")
neurosphere = c("E054", "E053")
thymus = c("E112", "E093")
adult_brain = c("E071", "E074", "E068", "E069", "E072", "E067", "E073", "E070")
adipose = c("E063")
adult_muscle = c("E100", "E108","E107")
adult_heart = c("E104", "E095", "E105", "E065")
smooth_muscle = c("E078", "E076", "E103", "E111")
adult_digestive = c("E109", "E106", "E075", "E101", "E102", "E110", "E077", "E079", "E094")

fetal = c("E080", "E081", "E082", "E083", "E084", "E085", "E086", "E088", "E089", "E090", "E092", "E093", "E017") # E091, placenta not tested because it is not strictly fetal
fetal_brain = c("E082", "E081")
ES_derived_neuronal = c("E007", "E009", "E010")

developmental = c(fetal, ES_derived_neuronal, ESC)

```


What proportion of the conserved elements are quiescent in all tissues?

```{r add chromHMM}
library(plyr)
conserved$region_id = paste0(conserved$chr, ":", conserved$start, "-", conserved$stop)

u = filter_with_bed(unaff_parent_variants, conserved)
u$region_id = get_region_id(u, conserved)

q = ddply(u, "region_id", function(df) data.frame(all_repressed = all(unlist(df[,grepl("E[0-1][0-9][0-9]_chromHMM", colnames(df))]) %in% c("9_Het", "13_ReprPC", "14_ReprPCWk", "15_Quies")), all_quiescent = all(unlist(df[,grepl("E[0-1][0-9][0-9]_chromHMM", colnames(df))]) %in% c("9_Het", "15_Quies")), all_polycomb = all(unlist(df[,grepl("E[0-1][0-9][0-9]_chromHMM", colnames(df))]) %in% c("13_ReprPC", "14_ReprPCWk")), n = nrow(df)))

```

Let's characterize these inactive elements more closely:

```{r characterize inactive}

quiescent = subset(conserved, region_id %in% q$region_id[q$all_quiescent])
polycomb_inactive = subset(conserved, (region_id %in% q$region_id[q$all_repressed]) & !(region_id %in% q$region_id[q$all_quiescent]))

active = subset(conserved, !(region_id %in% q$region_id[q$all_repressed]))
inactive = rbind(quiescent, polycomb_inactive)

vars_active = filter_with_bed(unaff_parent_variants, active)
vars_inactive = filter_with_bed(unaff_parent_variants, inactive)


vars_active$annotation = "Active CNEs"
vars_inactive$annotation = "Inactive CNEs"


v = rbind(vars_active, vars_inactive)

active_inactive = maps_adjust(v, v$annotation, maps_lm)
e = maps_ggplot(names(active_inactive$ps_adjusted), active_inactive$ps_adjusted, active_inactive$standard_error, already_ordered = FALSE, add_coding_fixed = FALSE, score_name = "")
e + ggtitle("CNEs inactive in all Roadmap Tissues\n")

# check phastcons score for CNEs (active v inactive)
library(phastCons100way.UCSC.hg19)

active_intervals = GRanges(seqnames=active$chr, IRanges(start = active$start, width = active$stop - active$start + 1))
inactive_intervals = GRanges(seqnames=inactive$chr, IRanges(start = inactive$start, width = inactive$stop - inactive$start + 1))

active_phastcons100 = scores(phastCons100way.UCSC.hg19, active_intervals)
inactive_phastcons100 = scores(phastCons100way.UCSC.hg19, inactive_intervals)

phastcons = data.frame(phastcons100 = c(active_phastcons100, inactive_phastcons100), set = c(rep("Active CNEs", length(active_phastcons100)), rep("Inactive CNEs", length(inactive_phastcons100))))

ggplot(phastcons, aes(set, phastcons100, fill = set)) + geom_violin() +
  theme_bw(base_size = 18) + 
  theme(strip.text = element_text(color="black"),strip.background = element_rect(fill="white", size=0),panel.border = element_blank()) + 
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(legend.title = element_blank()) + ggtitle("Active v. Inactive CNEs") + ylab("phastcons100\n") + xlab("") + guides(fill=FALSE)

```


What proportion of these elements are active in each tissue?

```{r active in each tissue}
library(ggplot2)

# TRUE means inactive and FALSE means active
a = ddply(u, "region_id", function(df) colSums(matrix(unlist(df[,grepl("E[0-1][0-9][0-9]_chromHMM", colnames(df))]) %in% c("9_Het", "13_ReprPC", "14_ReprPCWk", "15_Quies"), nrow = nrow(df))) == nrow(df))
colnames(a) = c("region_id", colnames(u)[grepl("E[0-1][0-9][0-9]_chromHMM", colnames(u))])

active_per_tissue = nrow(a) - colSums(a[-1])

df = data.frame(tissue = names(active_per_tissue), counts = active_per_tissue)
df$set = "Other"
df$set[df$tissue %in% paste0(developmental, "_chromHMM")] = "developmental"
df$set[df$tissue %in% c(paste0(blood, "_chromHMM"), paste0(hsc_and_b_cell, "_chromHMM"))] = "blood cells"
df$set[df$tissue %in% paste0(epithelial, "_chromHMM")] = "epithelial"
df$set[df$tissue %in% paste0(adult_brain, "_chromHMM")] = "adult_brain"
df$set[df$tissue %in% paste0(adult_muscle, "_chromHMM")] = "adult_muscle"
df$set[df$tissue %in% paste0(adult_heart, "_chromHMM")] = "adult_heart"
df$set[df$tissue %in% paste0(adult_digestive, "_chromHMM")] = "adult_digestive"

ordered_tissues = as.character(df$tissue[order(df$set)])
df$tissue = factor(df$tissue, levels = ordered_tissues, ordered = TRUE)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
ggplot(df) + geom_bar(aes(x = tissue, y = counts, fill = set), stat = "identity") + scale_fill_manual(values = cbbPalette) + 
  xlab("Tissues") + ylab("Number of Active CNEs (out of 4,300)") +
      theme_bw(base_size = 18) + 
      theme(strip.text = element_text(color="black"),strip.background = element_rect(fill="white", size=0),panel.border = element_blank()) + 
      theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
      theme(legend.title = element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())


```